Exploiting Intrinsic Noise for Heterogeneous Cell Control Under Time Delays and Model Uncertainties

### 2023-07-31 backup/paper_BrianJuly31.tex
Update text.

\documentclass[12pt]{iopart}
%\documentclass[12pt]{article}
%\usepackage{amsmath}
%\usepackage{cite}
\usepackage{amssymb,amsfonts}
%\usepackage{algorithmic}
\usepackage{graphicx}
%\usepackage{textcomp}
\usepackage{xcolor}
%\usepackage{graphicx}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\begin{document}

\def\brian[#1]{{\color{red} #1 }}
\def\michael[#1]{{\color{green} #1 }}

\title[Heterogeneous Control of Cell Differentiation Dynamics]{Exploiting Intrinsic Noise to Achieve Heterogeneous Control of Cell Differentiation Dynamics in the Presence of Time Delays, Model Uncertainties, and Extrinsic Variability}
\maketitle

@@

\begin{abstract}
Previous research has focused on robust control performance in the presence of noise, but understanding controllers that exploit noise remains incomplete. 
Motivated by Maxwells-Demon, we previously proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals. We found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
This work extends that analysis to include stochastic systems that trace a moving target, are affected by time delay, intrinsic noise, or measurement uncertainty.
 We find that noise-exploiting controllers can remain highly effective despite coarse approximations to the model's scale or incorrect estimations of key model parameters, and these controllers can even retain performance for significant time delays.  Together, these findings suggest that noise-exploiting control should be possible where models are always approximate, where parameters are always uncertain, and where observations are corrupted by errors.

@@

\section{Introduction}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{Cartoons.pdf}
\caption{{\bf Optogenetic control of multiple cells using a single light input.}
({\bf A}) Schematic of the light activated genetic system with autoregulation.
({\bf B}) Diagram of the stochastic SIMO control problem using two optogenetic cells sharing an input.
({\bf C}) Adapted from (PAPER). Noise exploiting controllers were optimized to a fully aware control input (I), and partially aware control input(III). Marginal distributions are solved using the finite state projection and show a break in symmetry using both controllers (II and IV) }
\label{cartoons}
\end{center}
\vspace{-0.3in}
\end{figure}
% XXX - Please modify figure as follows: (1) switch panels A and B to match caption. (2) Change title of panel B (previously panel A) to "External Observer, controller and actuator".  (3) Change title of panel A (previously panel B) to "Internal gene regulation pathway". (4) Change title of panel C to "Feedback control law and resulting performance"
% MMM - Made the edit

Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and is a frequent concern when attempting to control any human-made system such as vehicles\cite{XXX}, production machinery\cite{XXX}, or chemical processes\cite{XXX}.
Noise arrises from many sources, occasionally from quantum events and thermally-induced fluctuations but more commonly from unknown or uncharacterized physical processes, and as such, these fluctuations usually cannot be efficiently reproduced or predicted using purely physical or mechanistic models.
Rather, statistical models are usually employed where key mechanisms or dynamics are subject to stochastic variations or inputs that act as a proxy for random or un-modeled fluctuations elsewhere in the system.
In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \cite{XXX}.
For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem)\ref{XXX}, and most current approaches seek to control the system to minimize variations that result from the noisy inputs.

Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation\cite{XXX} or mRNA translation regulation \cite{XXX} or to modify these processes for practical use\cite{XXX}.
However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, additive noise model is far less realistic.
In this case, Brownian motion, discrete stochastic gene regulation, and noisy mRNA dynamics collectively generate a fundamentally stocastic environment that cells must effectively manage. 
At this scale, the order or timing of a single reaction event (e.g., the binding of a transcription factor to a promoter) can have dramatic consequences that could last for several cellular generations (e.g., the activation or repression of a gene that promotes unfettered cell growth and differentiation). 
The cell's drive towards homeostasis requires dealing with the inherently chaotic and noisy processes that reside within it, and despite these challenges, cells generally demonstrate strong capability to survive these noisy processes. 
When seeking to understand how such systems evolve or react, the central limit theorem and the Gaussian noise may not apply, and a more detailed statistical analysis of probabilistic behavior is needed uncover hidden properties of cellular control mechanisms.

One emerging field that is particularly dependent on the integration of control and noise is synthetic biology, which aims to develop modular\cite{Ng2019} and orthogonal \cite{Liu2018} components to sense and manipulate \cite{Sheets2020} complex logical systems, enabling them to exhibit a wide range of advanced biological behaviors\cite{Shin2020}. 
Advances in optogenetics have enhanced the ability to reliably actuate embedded systems within cells, offering the potential to exert precise temporal and spatial control on cellular components\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
These developments have facilitated computer-programmable regulation of cellular protein production through external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}. 
These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} . 
Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} to control synthetic systems to different stable points. 
% XXX Missing Citation above.

While most previous control efforts focussed on deterministic ordinary differential equation (ODE) formulations of the control problem, a few pioneering theoretical works have extended these analysis to include noise and have used stochastic simulations to show how systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}. 
It has recently been shown that new control techniques that leverage the complete probability distribution information of the system could actually harness the noise of single-cell gene regulation to achieve more complicated control objectives. 
For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins. 
In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{XXX} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
We then proposed the addition of a positive auto-regulation (Figure \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.

%The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states. For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd. In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.

This paper extends the analysis of the SIMO multicellular control problem by examining the impact of model uncertainties and fluctuating control objectives on the control performance. 
These uncertainties include coarse-grained approximations of the system dynamics, errors or extrinsic variations in the system parameter, and time delays between the observation of the cellular dynamics and actuation of the control process. 
In the following `Methods' we introduce our formulation of the chemical master equation to analyze the discrete stochastic distribution of cellular responses; we define multiple controllers and demonstrate the computation of their effects on cell dynamics; and we show how the control law can be optimized to improve performance.
In the `Results' section, we explore the how model approximations, parameter inaccuracies, and time delays affect control performance, and we demonstrate a simple scheme for controlling cells to track a dynamically changing reference signal. 
Using discrete stochastic models based on the chemical master equation, we demonstrate that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to follow different desired trajectories at different frequencies and phases.

%Figure\ \ref{cartoons}C(left) presents heat maps for two control laws, which a single external UV excitation signal based on observed cellular state.  Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions. Effective control was possible with both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)), even when the control model was an approximate simplification of the actual process dynamics.

\section{Methods}
In May et al. \cite{May2021}, we developed two models for the description of an optogenetically controlled gene expression system.
These first model consisted of six species to describe the light-activated association of two T7 split domains (species 1 and 2) which combine to form an active T7 polymerase (species 3) under optogenetic excitation. 
The active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active allele (species 5) that could then transcribe and translated to produce the desired protein product (species 6). 
The second, much simpler, single-species model was developed by assuming quasi-steady equilibrium for the first five species. 
Both models were independently parameterized using the same experimental data from Baumschlager et al. \cite{XXX}. 
Furthermore, an extension was made to each model to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
Through simulations, we showed that a feedback control law could be designed to force multiple cells to different and individually chosen equilibrium states using a single optogenetic control signal. 
We also showed that when this control law was parametrized using the simple model, it could be used effectively to control the behavior of the more complicated system, thus demonstrating that control performance could remain high despite inaccuracies in the model.
In the current work, we adopt the simpler of the two models and extend our analyses to consider the effects of additional model inaccuracies, including coarse-grained model approximations, parameter errors, and time delays, and we also explore the possibility that a single optogenetic control signal could drive the system to track temporally-changing reference signals.
Section \ref{sec:Model} introduces the model; Section \ref{sec:Stochastic} develops a Master Equation description of the models probabilistic dynamics; Section \ref{sec:Quantification} defines a control objective and optimizes that metric to obtain a baseline control law; and finally Sections \ref{sec:Scaling} and \ref{sec:Delay} introduces uncertainties into the model related t the granularity of the model approximation or the introduction of time delays, respectively.

\subsection{Model}\label{sec:Model}
To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al. \cite{May2021} (Figure. \ref{cartoons}A). 
This model comprises two reactions for production and degradation of the key protein. 
The nonlinear, UV-dependent production is defined by the following equation:
\begin{equation}
\nu_1(x,t)= \kappa\frac{x^\eta}{x^\eta+\beta^\eta}+ k_0 + u(UV(t)),
\label{prodRate}
\end{equation}
%
%These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
%\begin{equation}
%\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
%\label{rateeq}
%\end{equation}
%
where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches its half maximal strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the leakage rate from both promoters; and $u(UV(t))$ is the UV-dependent strength of the T7 promoter. 
Feedback enables the external modulation of the light input using the state of the system, thereby controlling the T7 promoter strength as a function of state rather than time and eliminating the explicit time dependance in $u(UV(t))$.
Protein degradation is assumed to be a first order process with rate $\gamma$:
\begin{equation}
\nu_2(x) = \gamma x.
\end{equation}
All baseline parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$ are presented in Table 1~\cite{May2021} and are fixed throughout the current study.

\begin{table}[]
\caption{Baseline Model Parameters}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter & Value   & units                     \\ \hline \hline
$\kappa$         & 0.406   & 1/min \\ \hline
$\beta$          & 20.0    & Molecules            \\  \hline
$\eta $           & 8.00    & unitless          \\  \hline
$k_0$           & 0.0001  & 1/min \\  \hline
$\gamma$         & 0.0203  & 1/min         \\  \hline
$u$              & dynamic & 1/min \\ \hline
\end{tabular}
\label{table}
\end{center}
\vspace{-0.2in}
\end{table}

\subsection{Stochastic analyses of the model}\label{sec:Stochastic}
To describe the discrete stochastic behavior of the above model for a population system with $N_{\rm c}$ cells, we define the current state of the system as the tuple of the non-negative numbers of proteins in each cell: $\mathbf{X}_i = [x_{1},x_{2},\ldots,x_{N_c}]_i\in \mathbb{Z}_{\ge 0}$, where the index $i$ denotes the enumeration of the state within the countably infinite set of all possible states, $\mathbf{X}_i\in\mathcal{X}=\{\mathbf{X}_1,\mathbf{X}_2,\ldots\}$.
The stoichiometry vector, $\mathbf{s}_{\mu}$, for reaction number $\mu$ is then defined as the change in state following that reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\mu$). Specifically, the $2N_c$ possible reactions are defined in pairs corresponding to production ($\mu\in\{1,3,5,\ldots\}$) and degradation ($\mu\in\{2,4,6,\ldots\}$) as:
\begin{eqnarray}
\mathbf{s}_1 &= \mathbf{e}_1,\ \mathbf{s}_2 = -\mathbf{e}_1,\nonumber\\ 
\mathbf{s}_3 &= \mathbf{e}_2,\ \mathbf{s}_4 = -\mathbf{e}_2,\nonumber\\
&\vdots \nonumber\\ 
\mathbf{s}_{\rm 2N_c-1} &= \mathbf{e}_{\rm N_c},\ \mathbf{s}_{\rm 2N_c} = -\mathbf{e}_{\rm N_c}, \label{Stochs}
 \end{eqnarray}
where each $\mathbf{e}_{i}\in\mathbb{Z}^{{\rm N_ c}}$ is is a Euclidean vector (i.e., unity for the $i^{\rm th}$ entry and otherwise zero). The corresponding propensity functions are:
\begin{eqnarray}
w_1(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_1^\eta}{x_1^{\eta}+\beta^{\eta}} + k_0,\ w_2(\mathbf{X}) = \gamma x_1,\nonumber\\
w_3(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_2^\eta}{x_2^{\eta}+\beta^{\eta}} + k_0,\ w_4(\mathbf{X}) = \gamma x_2,\nonumber\\
&\vdots \nonumber\\
w_{\rm 2N_c-1}(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_N^\eta}{x_N^{\eta}+\beta^{\eta}} + k_0,\ w_{\rm 2N_c}(\mathbf{X}) = \gamma x_N.\label{Props}
 \end{eqnarray}

These definitions of the stoichiometry and propensity functions allow us to implement the Gillespie stochastic simulation algorithm (SSA) \cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process. At each step two random numbers are generated to determine the time and the type of the next reaction. Given the current state $\mathbf{X}$, the time until the next reaction is distributed according to an exponential random with rate parameter equal to the inverse of the sum of the propensity functions:
\begin{equation}
Pr(\delta t = \tau) = \sum w_i(\mathbf{X}) \exp\left(- \sum{\mu=1}^{2N_c-1} w_\mu(\mathbf{X})\tau\right),
\end{equation}
and an instance of this random variable, $\delta t$,  can be sampled from this distribution using the expression:
\begin{equation}
\delta t = -\log\left(r_1 \sum_{\mu=1}^{2N_c-1}  w_\mu(\mathbf{X})\right),
\end{equation}
where $r_1$ is a uniform random variable between zero and one.
The probability of for the specific individual reaction $R_k$ to fire from all possible reactions is given by:
\begin{equation}
Pr(R_k)=\frac{w_k}{\sum w_{\mu=1}^{2N_c-1} (\mathbf{X})},
\end{equation}
and thus the reaction that fires at time $t+\tau$ is a categorical random variable pulled from $Pr(R_k)$.  The SSA is then simulated by stepping through time one reaction at a time and updating the state of the system by adding the stoichiometry of the reaction to the state of the system.

>>>>

\documentclass[12pt]{iopart}
%\documentclass[12pt]{article}
%\usepackage{amsmath}
%\usepackage{cite}
\usepackage{amssymb,amsfonts}
%\usepackage{algorithmic}
\usepackage{graphicx}
%\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
%\usepackage{graphicx}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}

\begin{document}
\def\brian[#1]{{\color{red} #1 }}
\def\michael[#1]{{\color{green} #1 }}

\title[Heterogeneous Control of Cell Differentiation Dynamics]{Exploiting Intrinsic Noise for Heterogeneous Cell Control Under Time Delays and Model Uncertainties}
\maketitle

@@

\begin{abstract}
Previous research has focused to enable robust control performance despite the presence of noise, but understanding how controllers may exploit that noise remains incomplete. 
Motivated by Maxwells-Demon, we previously proposed a cellular control regime in which the exploitation of stochastic noise can break symmetry between and allow for specific control of multiple cells using a single input signal (i.e., single-input-multiple-output or SIMO control).
The current work extends that analysis to include uncertain stochastic systems where system dynamics are are affected by time delays, intrinsic noises, and model uncertainty. 
We find that noise-exploiting controllers can remain highly effective despite coarse approximations to the model's scale or incorrect estimations or extrinsic noise in key model parameters, and these controllers can even retain performance under substantial observer or actuator time delays.  
We also demonstrate how SIMO controllers could drive multi-cell systems to follow different trajectories with different phases and frequencies.
Together, these findings suggest that noise-exploiting control should be possible even in the practical case where models are aways approximate, where parameters are always uncertain, and where observations are corrupted by errors.

@@

\section{Introduction}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{Cartoons.pdf}
\caption{{\bf Single-input-multiple-output (SIMO) control of multiple cells using a single optogenetic input.}
({\bf A}) Schematic of the light-activated genetic system with auto-regulation.
({\bf B}) Diagram of the stochastic SIMO control problem using two optogenetic cells sharing a single input.
({\bf C}) Noise exploiting controllers were optimized to define a fully aware control input (I) and a partially aware control input (III), where the control signal (color scale) depends on the observed expression within the cell or cells (x- and y-axes).  (II and IV) Corresponding steady state marginal distributions for the different cells (red and blue) under these controllers demonstrate a clear a break in symmetry. Dashed lines represent the control target objective. Control performance (RMSE) is show above each distribution.}
\label{cartoons}
\end{center}
\vspace{-0.3in}
\end{figure}

Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and is a frequent concern when attempting to control any human-made system such as vehicles \brian[\cite{XXX}], production machinery \brian[\cite{XXX}], or chemical processes \brian[\cite{XXX}].
\brian[Please look for some relevant references or a review of control under noise. I highlight several missing refs in red.]
Noise arrises from many sources, occasionally from quantum events and thermally-induced fluctuations, but more commonly from unknown or uncharacterized physical processes, and as such, these fluctuations usually cannot be efficiently reproduced or predicted using purely physical or mechanistic models.
Rather, statistical models are usually employed where key mechanisms or dynamics are subject to stochastic variations or inputs that act as a proxy for random or un-modeled fluctuations elsewhere in the system.
In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \brian[\cite{XXX}].
For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem) \brian[\cite{XXX}], and most current approaches seek to control the system to minimize variations that result from the noisy inputs.

Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation \brian[\cite{XXX}] or mRNA translation regulation \brian[\cite{XXX}]or to modify these processes for practical use \brian[\cite{XXX}].
However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, the additive noise model is much less realistic.
In this case, Brownian motion, discrete stochastic gene regulation, and noisy mRNA dynamics collectively generate a fundamentally stochastic environment that cells must effectively manage. 
At this scale, the order or timing of a single reaction event (e.g., the binding of a transcription factor to a promoter) can have dramatic consequences that could last for several cellular generations (e.g., the activation or repression of a gene that promotes unfettered cell growth and differentiation). 
The cell's drive towards homeostasis requires dealing with the inherently chaotic and noisy processes that reside within it, and despite these challenges, cells generally demonstrate strong capability to survive these noisy processes. 
When seeking to understand how such systems evolve or react, the central limit theorem and the Gaussian noise may not apply, and a more detailed statistical analysis of probabilistic behavior is needed uncover hidden properties of cellular control mechanisms.

One emerging field that is particularly dependent on the integration of control and noise is synthetic biology, which aims to develop modular\cite{Ng2019} and orthogonal \cite{Liu2018} components to sense and manipulate \cite{Sheets2020} complex logical systems, enabling them to exhibit a wide range of advanced biological behaviors\cite{Shin2020}. 
Advances in optogenetics have enhanced the ability to reliably actuate embedded systems within cells, offering the potential to exert precise temporal and spatial control on cellular components\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
These developments have facilitated computer-programmable regulation of cellular protein production through external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}. 
These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} . 
Classical and modern control methods like PID control and model predictive control have been implemented in such systems \brian[\cite{XXX}] to control synthetic systems to different stable points.

While most previous control efforts focussed on deterministic ordinary differential equation (ODE) formulations of the control problem, a few pioneering theoretical works have extended these analysis to include noise and have used stochastic simulations to show how systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}. 
It has recently been shown that new control techniques that leverage the complete probability distribution information of the system could actually harness the noise of single-cell gene regulation to achieve more complicated control objectives. 
For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins. 
In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al. \brian[\cite{XXX}] for the expression from a transcription promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
We then proposed the addition of a positive auto-regulation (Figure \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.

This paper extends the analysis of the SIMO multicellular control problem by examining the impact of model uncertainties and fluctuating control objectives on the control performance. 
These uncertainties include coarse-grained approximations of the system dynamics, errors or extrinsic variations in the system parameter, and time delays between the observation of the cellular dynamics and actuation of the control process. 
In the following `Methods' we introduce our formulation of the chemical master equation (CME) to analyze the discrete stochastic distribution of cellular responses; we define multiple controllers and demonstrate the computation of their effects on cell dynamics; and we show how the control law can be optimized to improve performance.
In the `Results' section, we explore the how model approximations, parameter inaccuracies, and time delays affect control performance, and we demonstrate a simple scheme for controlling cells to track a dynamically changing reference signal. 
Using discrete stochastic models based on the chemical master equation, we demonstrate that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to follow different desired trajectories at different frequencies and phases.

\section{Methods}
In May et al. \cite{May2021}, we developed two models for the description of an optogenetically controlled gene expression system.
These first model consisted of six species to describe the light-activated association of two T7 split domains (species 1 and 2) which combine to form an active T7 polymerase (species 3) under optogenetic excitation. 
The active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active allele (species 5) that could then transcribe and translated to produce the desired protein product (species 6). 
The second, much simpler, single-species model was developed by assuming quasi-steady equilibrium for the first five species. 
Both models were independently parameterized using the same experimental data from Baumschlager et al. \brian[\cite{XXX}]. 
Furthermore, an extension was made to each model to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
Through simulations, we showed that a feedback control law could be designed to force multiple cells to different and individually chosen equilibrium states using a single optogenetic control signal. 
We also showed that when this control law was parametrized using the simple model, it could be used effectively to control the behavior of the more complicated system, thus demonstrating that control performance could remain high despite inaccuracies in the model.
In the current work, we adopt the simpler of the two models and extend our analyses to consider the effects of additional model inaccuracies, including coarse-grained model approximations, parameter errors, and time delays, and we also explore the possibility that a single optogenetic control signal could drive the system to track temporally-changing reference signals.
Section \ref{sec:Model} introduces the model; Section \ref{sec:Stochastic} develops a Master Equation description of the models probabilistic dynamics; Section \ref{sec:Quantification} defines a control objective and optimizes that metric to obtain a baseline control law; and Sections \ref{sec:Scaling} and \ref{sec:Delay} introduce uncertainties into the model related to the granularity of the model approximation or the introduction of time delays, respectively.

\subsection{Model}\label{sec:Model}
To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al. \cite{May2021} (Figure. \ref{cartoons}A). 
This model comprises two reactions for production and degradation of the key protein. 
The nonlinear, UV-dependent production is defined by the following equation:
\begin{equation}
\nu_1(x,t)= \kappa\frac{x^\eta}{x^\eta+\beta^\eta}+ k_0 + u(UV(t)),
\label{prodRate}
\end{equation}
where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches its half maximal strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the leakage rate from both promoters; and $u(UV(t))$ is the UV-dependent strength of the T7 promoter. 
Feedback enables the external modulation of the light input using the state of the system, thereby controlling the T7 promoter strength as a function of state rather than time and eliminating the explicit time dependance in $u(UV(t))$.
Protein degradation is assumed to be a first order process with rate $\gamma$:
\begin{equation}
\nu_2(x) = \gamma x.
\end{equation}
All baseline parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$ are presented in Table 1~\cite{May2021} and are fixed throughout the current study.

\begin{table}[]
\caption{Baseline Model Parameters}
\begin{center}
\begin{tabular}{|c|c|c|l|}
\hline
Parameter & Value   & Units & Meaning                     \\ \hline \hline
$\kappa$         & 0.406   & 1/min & maximal production rate \\ \hline
$\beta$          & 20.0    & molecules  & concentration at half-max activation        \\  \hline
$\eta $           & 8.00    & unitless  & cooperativity factor        \\  \hline
$k_0$           & 0.0001  & 1/min & promote leakage rate \\  \hline
$\gamma$         & 0.0203  & 1/min & degradation/dilution rate         \\  \hline
$u(t)$              & variable & 1/min & applied control signal  \\ \hline
\end{tabular}
\label{table}
\end{center}
\vspace{-0.2in}
\end{table}

\subsection{Stochastic analyses of the model}\label{sec:Stochastic}
To describe the discrete stochastic behavior of the above model for a population of $N_{\rm c}$ cells, we define the current state of the system as the tuple of the non-negative numbers of proteins in each cell: $\mathbf{X}_i = [x_{1},x_{2},\ldots,x_{N_c}]_i\in \mathbb{Z}_{\ge 0}$, where the index $i$ denotes the enumeration of the state within the countably infinite set of all possible states, i.e., $\mathbf{X}_i\in\mathcal{X}=\{\mathbf{X}_1,\mathbf{X}_2,\ldots\}$.
The stoichiometry vector, $\mathbf{s}_{\mu}$, for reaction number $\mu$ is then defined as the change in state following that reaction event (e.g., $\mathbf{X}_i \rightarrow \mathbf{X}_i + \mathbf{s}_\mu$). Specifically, the $2N_c$ possible reactions are defined in pairs corresponding to production ($\mu\in\{1,3,5,\ldots\}$) and degradation ($\mu\in\{2,4,6,\ldots\}$) as:
\begin{eqnarray}
\mathbf{s}_1 &= \mathbf{e}_1,\ \mathbf{s}_2 = -\mathbf{e}_1,\nonumber\\ 
\mathbf{s}_3 &= \mathbf{e}_2,\ \mathbf{s}_4 = -\mathbf{e}_2,\nonumber\\
&\vdots \nonumber\\ 
\mathbf{s}_{\rm 2N_c-1} &= \mathbf{e}_{\rm N_c},\ \mathbf{s}_{\rm 2N_c} = -\mathbf{e}_{\rm N_c}, \label{Stochs}
 \end{eqnarray}
where each $\mathbf{e}_{i}\in\mathbb{Z}^{{\rm N_ c}}$ is is a Euclidean vector (i.e., unity for the $i^{\rm th}$ entry and otherwise zero). The corresponding propensity functions are:
\begin{eqnarray}
w_1(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_1^\eta}{x_1^{\eta}+\beta^{\eta}} + k_0,\ w_2(\mathbf{X}) = \gamma x_1,\nonumber\\
w_3(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_2^\eta}{x_2^{\eta}+\beta^{\eta}} + k_0,\ w_4(\mathbf{X}) = \gamma x_2,\nonumber\\
&\vdots \nonumber\\
w_{\rm 2N_c-1}(\mathbf{X}) &= u(\mathbf{X},t)  + \kappa \frac{x_N^\eta}{x_N^{\eta}+\beta^{\eta}} + k_0,\ w_{\rm 2N_c}(\mathbf{X}) = \gamma x_N.\label{Props}
 \end{eqnarray}

These definitions of the stoichiometry and propensity functions allow us to implement the Gillespie stochastic simulation algorithm (SSA) \cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process. At each step, two random numbers are generated to determine the time and the type of the next reaction. Given the current state $\mathbf{X}$, the time until the next reaction is distributed according to an exponential random with rate parameter equal to the inverse of the sum of the propensity functions:
\begin{equation}
Pr(\delta t = \tau) = \sum w_i(\mathbf{X}) \exp\left(- \sum{\mu=1}^{2N_c-1} w_\mu(\mathbf{X})\tau\right),
\end{equation}
and an instance of this random variable, $\delta t$,  can be sampled from this distribution using the expression:
\begin{equation}
\delta t = -\log\left(r_1 \sum_{\mu=1}^{2N_c-1}  w_\mu(\mathbf{X})\right),
\end{equation}
where $r_1$ is a uniform random variable between zero and one.
The probability for the specific individual reaction $R_k$ to fire from all possible reactions is given by:
\begin{equation}
Pr(R_k)=\frac{w_k}{\sum w_{\mu=1}^{2N_c-1} (\mathbf{X})},
\end{equation}
and which specific reaction that fires at time $t+\tau$ is a categorical random variable pulled from $Pr(R_k)$.  The SSA is then simulated by stepping through time one reaction at a time and updating the state of the system by adding the stoichiometry of the reaction to the state of the system.

However, a more direct analysis of the chemical master equation (CME) is necessary to quantify performance and optimize the controller design. The high dimensional CME is a linear ODE that describes the time-dependent changes in probability mass of all possible states. Using the specified reaction propensities and stoichiometries, the CME can be expressed as:
\begin{equation}
\frac{d}{dt}P(\mathbf{X}_i)=\sum_{\mu=1}^{2{\rm N_c}}
\left(
-w_{\mu}(\mathbf{X}_i)P(\mathbf{X}_i)
+w_{\mu}(\mathbf{X}_i-\mathbf{s}_{\mu})P(\mathbf{X}_i-\mathbf{s}_{\mu})
\right).\label{CMEindex}
\end{equation}
For convenience, the CME can also be formulated more compactly in matrix-vector form as:
\begin{equation}
\frac{d}{dt}\mathbf{P}=(\mathbf{A}_0+\textbf{Bu}^{\mathcal{C}})\mathbf{P},\label{CME}
\end{equation}
where $\mathbf{P} = [P(\mathbf{X}_1), P(\mathbf{X}_2), \ldots ]^T$ is the enumerated probability mass vector for all possible states of the system; $\mathbf{A}_0$ is the infinitesimal generator of the stochastic process due to the autoregulation promoter and degradation events;  $\textbf{u}^{\mathcal{C}} =[u^{\mathcal{C}}(\mathbf{X}_1), u^{\mathcal{C}}(\mathbf{X}_2), \ldots ]^T$ is the collection of control inputs associated with each state; and $\textbf{Bu}^{\mathcal{C}}$ is the contribution that these control inputs make to the infinitesimal generator when included into the feedback process.
